\documentclass[12pt]{article}
\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{titlesec}
\usepackage{titling}
\usepackage{graphicx}
\usepackage{url}
\usepackage{amsmath}
\usepackage{color}
\usepackage[colorlinks=true, urlcolor=blue, linkcolor=blue, citecolor=blue]{hyperref}

% Title formatting courtesy of Damein Koon
\pretitle{\LARGE\bfseries}
\posttitle{\par\vspace{2em}}
\preauthor{\large}
\postauthor{\par\vspace{1em}}
\predate{\large}
\postdate{\par\vspace{3em}}

% Corrected title hooks for vertical and horizontal centering
\renewcommand\maketitlehooka{\null\vfill\begin{center}}
\renewcommand\maketitlehookd{\end{center}\vfill\null}

\title{Supermassive Black Hole Formation Information}
\author{Krupa Pothiwala \& Liz Hayes \\[0.5em]
February 2025}
\date{}

% Comment commands
\newcommand{\com}[2]{\textcolor{#1}{[#1: #2]}}
\newcommand{\bcom}[2]{\textcolor{#1}{\textbf{[#1: #2]}}}
\definecolor{LH}{rgb}{1,0,0}
\definecolor{KP}{rgb}{0,0.0,0.99}

\begin{document}
\maketitle

\newpage

\com{KP}{I was thinking about we can research what we think fits and whichever one we like we can talk about? As long as it doesn't overlap.}
\section{Background Information:}


\begin{enumerate}
    \item Supermassive black holes are classically defined as black holes with a mass above $100{,}000\,M_\odot$ and some have masses of several billion $M_\odot$.
    \item We see supermassive black holes (millions to billions of times the Sun’s mass) very early in the universe; only about 700 million years after the Big Bang.
    \item As far as we know all the galaxies have a supermassive Black Holes at the center of them.
    \item Black hole of mass greater than $5$ billion $M_\odot$ are considered ultramassive black holes (UMBHs or UBHs).
\end{enumerate}

\section{Sources}

\subsection*{Source 1 - Formation of the First Supermassive Black Holes}
\url{https://arxiv.org/abs/astro-ph/0212400} 
\begin{itemize}

\item Quasars at redshift $z > 6$ show that billion-solar-mass black holes already existed when the universe was less than one billion years old. If black holes formed only from normal stars and then slowly grew by accreting gas, there may not have been enough time for them to reach such large masses. Because of this time problem, the paper investigates whether a faster formation pathway existed — specifically, whether massive gas clouds in the first galaxies could collapse directly into large black hole “seeds” instead of starting from small stellar remnants.

\item Direct Collapse Scenario:
\begin{itemize}
    \item The paper studies early dwarf galaxies with virial temperatures around or above 10,000 K. The gas in these galaxies is assumed to be metal-free.
    \item A background of UV radiation is present in the early universe. This UV radiation destroys molecular hydrogen (H$_2$). Without H$_2$, the gas cannot cool efficiently below about 10,000 K.
    \item Because the gas stays hot, it does not fragment into many small stars. Instead, a large amount of gas collapses into one (or sometimes two) massive central clumps. These massive clumps can later collapse into a supermassive black hole.
    \item A “seed” is the initial, smaller black hole that later grows into a supermassive black hole by accreting more matter.
    \item Virial temperature is the temperature gas reaches when gravity and pressure balance in a collapsing halo.
\end{itemize}

\item What they did:
\begin{itemize}
    \item The authors simulated the collapse of an early galaxy-sized dark matter halo with a mass of about $10^8$ solar masses at redshift $z \approx 10$. They used a numerical method called Smoothed Particle Hydrodynamics (SPH) to model how the gas moves, cools, and collapses. They compared different scenarios: one where molecular hydrogen (H$_2$) cooling was allowed and one where it was suppressed. They also included the effects of ultraviolet (UV) background radiation and tested how different amounts of galaxy rotation (spin) affected the collapse.
\end{itemize}

\item When H$_2$ Suppressed:
\begin{itemize}
    \item When H$_2$ cooling is suppressed, the gas stays hot at about 10,000 K and can only cool through atomic hydrogen; meaning the gas loses energy when excited hydrogen atoms emit radiation (like Lyman-alpha light), allowing the gas temperature to decrease.
    \item Because it does not cool very much, the gas cannot break apart into many small pieces to form stars. Instead, it collapses almost uniformly toward the center of the galaxy.
    \item This forms a very massive central clump of gas about one million solar masses within a region smaller than one parsec.
    \item In some cases, two large clumps form.
    \item Overall, about 10 percent or more of the galaxy’s gas ends up concentrated in the center.
\end{itemize}

\item When H$_2$ is not suppressed:
\begin{itemize}
    \item When H$_2$ cooling is allowed, the gas cools much more efficiently, down to about 200 K.
    \item This lower temperature reduces pressure and allows the gas to fragment into many small clumps.
    \item Stars form early, and supernova explosions enrich the gas with metals, making cooling even easier.
    \item Because the gas fragments and forms stars instead of collapsing as one large object, there is no massive central collapse and no direct formation of a supermassive black hole.
\end{itemize}

\item Final Evolution:
\begin{itemize}
    \item After the massive central clump forms, it continues collapsing under gravity and becomes extremely dense.
    \item As it contracts, it eventually becomes optically thick, meaning radiation can no longer escape easily.
    \item At this stage, the gas likely forms a supermassive star with a mass of about one million solar masses.
    \item Because such a star is unstable under its own gravity, relativistic effects eventually cause it to collapse.
    \item Most of its mass (about 90 percent) ends up forming a massive black hole seed, which can later grow into a supermassive black hole.
    \item “Relativistic effects” meaning that when the supermassive star becomes extremely massive and dense, corrections from Einstein’s theory of general relativity become important, making gravity stronger than predicted by normal (Newtonian) physics and causing the star to become unstable and collapse into a black hole.
\end{itemize}

\item Conditions Required for Direct Collapse:
\begin{itemize}
    \item The gas must be metal-free and exposed to strong ultraviolet (Lyman–Werner) radiation, which suppresses molecular hydrogen (H$_2$) formation.
    \item Without H$_2$, the gas cannot cool efficiently or fragment into stars.
    \item The dark matter halo must also be massive (around $10^8$ solar masses or more) and have a virial temperature above about 10,000 K so that atomic hydrogen cooling dominates.
\end{itemize}

\item Conclusions:
\begin{itemize}
    \item Direct collapse is physically possible in early galaxies under the conditions listed above.
    \item This process can create large black hole seeds of about $10^5$ to $10^6$ solar masses, which helps explain how quasars formed so early in the universe.
    \item However, only a small fraction of early halos likely had the exact conditions needed for this to occur.
    \item Quasars are extremely bright objects powered by gas falling into supermassive black holes, and they show us that very massive black holes already existed in the early universe.
    \item Halos are large dark matter structures that gravity pulls gas into, forming the first galaxies. They are the environments where supermassive black holes could have formed.
\end{itemize}

\end{itemize}

\subsection*{Source 2 - Formation of Supermassive Black Holes by Direct Collapse in Pregalactic Halos}
\url{https://arxiv.org/abs/astro-ph/0602363}

\begin{itemize}

\item This paper proposes another way supermassive black holes could form very early in the universe. Instead of starting from massive stars, the authors investigate whether gas in early dark matter halos could collapse directly through gravitational instabilities and form a black hole that then grows rapidly.

\item Direct Collapse with Angular Momentum Transport:
\begin{itemize}
    \item The paper focuses on early dark matter halos with virial temperatures around or above 10,000 K.
    \item The gas is assumed to be metal-free and able to cool through atomic hydrogen (but not fragment easily).
    \item The main challenge is angular momentum — gas normally rotates, which prevents collapse.
    \item The authors propose a mechanism called “bars within bars.”
    \item In this process, large-scale gravitational instabilities form bar-like structures in the gas disk.
    \item These bars transport angular momentum outward.
    \item As angular momentum is removed, gas flows rapidly inward toward the center.
    \item Instead of forming many stars, gas builds up in the nucleus of the halo.
\end{itemize}

\item What They Did:
\begin{itemize}
    \item The authors developed an analytic (theoretical) model rather than a numerical simulation.
    \item They calculated when gas disks in early halos would become gravitationally unstable.
    \item They analyzed how efficiently angular momentum could be transported outward.
    \item They studied how gas collapses inward under these unstable conditions.
    \item They estimated the mass growth rate of the forming black hole.
    \item They considered how radiation pressure and energy feedback affect the collapse.
\end{itemize}

\item Formation of a “Quasistar”:
\begin{itemize}
    \item As gas rapidly flows inward, it forms a massive, pressure-supported object called a quasistar.
    \item A quasistar has:
    \begin{itemize}
        \item A radiation-pressure-dominated outer envelope
        \item A dense, gas-pressure core
    \end{itemize}
    \item Gas continues falling inward, squeezing the core.
    \item The core temperature steadily increases.
\end{itemize}

\item Core Collapse:
\begin{itemize}
    \item When the core temperature reaches about $10^9$ K, neutrino cooling becomes important.
    \item Neutrinos carry energy away very efficiently.
    \item The core loses pressure support and collapses.
    \item A small black hole forms, with a mass of about 10--20 $M_\odot$.
\end{itemize}

\item Rapid Black Hole Growth:
\begin{itemize}
    \item After forming, the black hole is surrounded by a huge gas envelope.
    \item It accretes gas from the quasistar.
    \item Growth can occur at a super-Eddington rate (faster than normal limits).
    \item The black hole can grow to about $10^4$--$10^6$ $M_\odot$ in a relatively short time.
    \item This provides a massive seed that can later grow into a billion-solar-mass black hole.
\end{itemize}

\item Conditions Required:
\begin{itemize}
    \item Dark matter halo virial temperature $\gtrsim 10{,}000$ K
    \item Metal-free or nearly metal-free gas
    \item Gas disk becomes gravitationally unstable
    \item Efficient angular momentum transport (“bars within bars”)
    \item Limited fragmentation into stars
\end{itemize}

\item Conclusions:
\begin{itemize}
    \item Direct collapse through gravitational instabilities is physically possible in early halos.
    \item Black holes may start small ($\sim 10$--$20\,M_\odot$) but grow extremely rapidly.
    \item This mechanism can produce black holes of $10^4$--$10^6\,M_\odot$ early in cosmic history.
    \item It helps explain how quasars existed at redshift $z > 6$.
    \item Only a fraction of halos (those with low spin and proper conditions) would undergo this process.
\end{itemize}

\end{itemize}

\subsection*{Source 3 - Supermassive black hole formation via collisions in black hole clusters}
\url{https://arxiv.org/abs/2406.13072}
\begin{itemize}
    \item The paper investigates how dense clusters of stellar-mass black holes, possibly formed at the centers of galaxies via mass segregation, can merge to produce massive black hole seeds in the early universe. The authors include the effect of an external potential (such as gas inflows from galaxy mergers) on the evolution of these clusters, which can enhance merger rates and cluster contraction. Results show that the efficiency of forming a massive central black hole correlates strongly with the ratio of cluster velocity dispersion to the speed of light, reaching up to $5–8\%$ efficiency in realistic systems. They conclude this channel is a feasible contributor to SMBH seed formation, potentially producing seeds of at least $\sim10^3 M\odot$, and should be included in models of black hole population statistics.
    \item Dense Cluster Formation:
    \begin{itemize}
        \item The local universe also hosts nuclear star clusters (NSCs) at their centers. There have been multiple cases where the NSCs and SMBH are known to co-exists.
        \item Due to fragmentation in molecular clouds, we can have lots of stars packed very close together. 
        \item Dense clusters of stellar-mass black holes can form via mass segregation and may undergo runaway mergers. forming a central intermediate-mass black hole (IMBH). With mass of $10^{2-4} M\odot$. 
        \item Runaway means that they feed on itself. It is easier to snowball because mergers create bigger black holes which causes more runaway mergers. 
        \item The gas in the molecular cloud would feed the IMBH seed and cause it to become a massive BH seed.
        \item Gas inflows into the cluster’s environment steepen the gravitational potential, increasing velocity dispersions and boosting merger efficiency.
        \item The formation efficiency of very massive black holes is principally determined by the ratio of cluster velocity dispersion to the speed of light. 
        \item This mechanism could produce intermediate-mass seeds ($\sim 10^3 - 10^4 M\odot$) and thus plays a significant role in early SMBH assembly.
        \item The other important part is played by dynamical friction, which can also cause massive objects to sink to the center of a star cluster, where at the end of their lifetime the massive objects will evolve into stellar mass BHs or neutron stars, forming a dark core.
        \item Dynamical friction is loss of momentum and kinetic energy of moving bodies through gravitational interactions with surrounding matter in space. \url{https://en.wikipedia.org/wiki/Dynamical_friction}
        \item Something the BHs can escape due to so many interactions however, gas inflows after galaxy mergers could steepen the potential of the cluster contributing to the BHs seeds.
    \end{itemize}
    \item Method
    \begin{itemize}
        \item The authors consider a cluster composed entirely of stellar-mass black holes located in a galactic nucleus.
        \item The cluster is initially supported against gravitational collapse by random motions of the black holes (velocity dispersion).
        \item In this state, the system is approximately virialized, meaning gravitational attraction is balanced by kinetic energy.
        \item When black holes interact gravitationally, three-body encounters can form binary black hole systems.
        \item Once binaries form, they can interact with other black holes in the cluster.
        \item These interactions make the binary tighter (“harder”) and transfer energy to surrounding black holes.
        \item This energy injection acts like heating the cluster. The heating prevents the cluster from collapsing too quickly.
        \item Each time a binary interacts with another black hole: the binary becomes more tightly bound, and excess energy is transferred to the third object. 
        \item The third black hole often receives a velocity kick. Sometimes that kick is strong enough that the black hole escapes the cluster. This process slowly removes mass from the cluster.
        \item When a binary becomes very tight, gravitational wave emission becomes important. Gravitational waves remove orbital energy from the binary. Eventually the two black holes merge into a single larger black hole. This increases the average black hole mass in the cluster.
        \item When the binaries merge the source of the heat disappears or lessen in the system meaning the system starts to collapse.
        \item During this phase: density rises dramatically, encounter rates increase, and mergers become much more frequent.
        \item A runaway growth process begins where one black hole grows rapidly through repeated mergers. This produces a massive central black hole.
    \end{itemize}
    \item Conclusion
    \begin{itemize}
        \item The paper concludes that dense clusters of stellar-mass black holes provide a viable pathway for forming intermediate-mass black hole seeds that can later grow into supermassive black holes.
        \item Gas accumulation in galactic centers deepens the gravitational potential, which helps prevent black hole ejections and makes runaway mergers more efficient. This means that black hole cluster mergers may be more effective in gas-rich galactic nuclei.
        \item The seeds produced in this model ($\sim 10^3–10^4$ $M\odot$) are significantly larger than stellar remnants. This helps explain how some supermassive black holes observed in the early universe could grow to billions of solar masses within a billion years.
        \item Because the process appears efficient under certain conditions, the authors argue that black hole cluster mergers should be considered alongside other seed formation models when studying the cosmic population of supermassive black holes.
    \end{itemize}
\end{itemize}

\com{KP}{If I have time but I think we should only do three total sources because of 15 min time.}
\subsection*{Source 4 - Super-Critical Growth of Massive Black Holes from Stellar-Mass Seeds} 
\url{https://arxiv.org/abs/1402.6995}
\begin{itemize}
    \item This paper proposes that SMBHs in the early universe could grow from stellar-mass seeds through super-Eddington accretion. The authors analyze accretion flows using the slim disk model, which is radiatively inefficient and allows black holes to accrete mass faster than the classical Eddington limit would suggest. In this regime, much of the radiation is trapped in the flow and advected into the black hole instead of escaping, which reduces radiative feedback. Because of the low radiative efficiency, the black hole mass can grow exponentially on a short timescale almost independent of spin. The authors conclude that intermittent super-critical accretion episodes could solve the problem of forming billion-solar-mass quasars within the first ~0.8 billion years of the universe.
    \com{KP}{I need to learn what radiative efficiency is}
    \item About the Acceretion disk
    \begin{itemize}
        \item 
    \end{itemize}
    \item The slim disk solution 
    \begin{itemize}
        \item 
    \end{itemize}
    \item Conclusion
    \begin{itemize}
        \item 
    \end{itemize}
\end{itemize}

\end{document}